---
title: "Logistic Regression Analyses"
author: "Cormac Monaghan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
toc: true
format:
  html:
    html-math-method: katex
    code-tools: true
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
execute:
  warning: false
---

```{r setup, include = FALSE}
library(dplyr)

source(here::here("./02__Models/00__Functions.R"))
```

# Introduction
This document performs multiple binary logistic regression analyses to assess how well several health related problems and health protective behaviors (DVs) are explained by procrastination (IV). This analysis uses data from the 2020 wave of the [Health and Retirement Study (HRS)](https://hrs.isr.umich.edu/).

```{r import-data, include = FALSE}
path_data <- "./01__Data/02__Processed/"

# Reading in data
health_data <- readxl::read_xlsx(here::here(path_data, "Health_HRS.xlsx"))
```

# Logistic Regression
Logistic regression is a statistical method used for predicting the probability of a binary outcome based on one or more predictor variables. Unlike linear regression, which predicts continuous values, logistic regression models the probability of the dependent variable belonging to a particular category.

The logistic regression equation can be written as:

$$
P(Y = 1 ∣ X) = \frac{1}{1 + e^{{\beta_{0} + \beta_{1}X_{1} + \beta_{2}X_{2} + \dots + \beta_{n}X_{n}}}}
$$
where $P(Y = 1 ∣ X)$ is the probability of the dependent variable ($Y$) being 1 given the values of predictors ($X$). $e$ is the base of the natural logarithm. $\beta_{0}, \beta_{1}, \beta_{2}, \dots \beta_{n}$ are the coefficients representing the effect of each predictor variable on the outcome and $X_{1}, X_{2}, \dots X_{n}$ are the values of the predictor variables.

## Defining health problems and health protection
Within the HRS dataset, we have a range of both health problems and health protective behaviors. These have been coded as either $0$ (not present/done) or $1$ (present/done). 

```{r view-data, echo = TRUE}
health_data %>%
  select(Back_pain:Alcohol, Smoker_current, Blood_pressure:Dental_visit_2_years,
         Total_procrastination) %>%
  str()
```

For the purpose of our analysis we will define these variables into 2 vectors: one for problems and one for protective behaviors
```{r vector-defining}
# Defining a vector of health problems
health_problems <- c("Back_pain", "Headache", "Fatigue", 
                     "Alcohol", "Smoker_current", "Blood_pressure", 
                     "Diabetes", "Cholesterol", "Heart_condition")

# Defining a vector of health protective behaviors
health_protection <- c("Prostate_exam", "Mammogram", "Cholesterol_screening",
                       "Pap_smear", "Flu_shot", "Dental_visit_2_years")
```

## Performing logistic analysis
For each variable we will apply a custom function from the [Functions.R](https://github.com/C-Monaghan/Procrastination_Health/blob/main/02__Models/00__Functions.R) script. This function (`logit_model()`) applies a `glm()` model with the `family = binomial(link = "logit")` argument, whilst also storing summary output, various fit metrics generated by the [DescTools](https://andrisignorell.github.io/DescTools/) package, and the relevant odds ratio for that variable.

```{r log-model}
# Fitting binary logistic regressions (with no covariate) ----------------------
# Fitting (health problems)
problem_models_base <- lapply(health_problems, logit_model,
                              predictor = "Total_procrastination",
                              data = health_data, type = "base")

# Fitting (health protection)
protection_models_base <- lapply(health_protection, logit_model,
                                 predictor = "Total_procrastination",
                                 data = health_data, type = "base")

# Fitting binary logistic regressions (with covariate) -------------------------
problem_models_control <- lapply(health_problems, logit_model, 
                                 predictor = "Total_procrastination", 
                                 data = health_data, type = "control")

protection_models_control <- lapply(health_protection, logit_model, 
                                    predictor = "Total_procrastination", 
                                    data = health_data, type = "control")
```

# Visualisation
To better summarize and visualise each model we can create individual HTML tables using the [sjPlot](https://strengejacke.github.io/sjPlot/) package and an odds ratio plot using the `log_odds_plot()` function, again from the [Functions.R](https://github.com/C-Monaghan/Procrastination_Health/blob/main/02__Models/00__Functions.R). 

```{r include = FALSE}
# Defining a vector of health problems
health_problems <- c(
  "Back Pain", "Headache", "Fatigue", "Drinking", "Smoking", 
  "Blood Pressure", "Diabetes", "Cholesterol", "Heart Condition")

# Defining a vector of health protective behaviors
health_protection <- c(
  "Prostate Exam", "Mammogram", "Cholesterol Screening",
  "Pap Smear", "Flu Shot", "Dental Visit (2yrs)")

# Extracting relevant GLM results ----------------------------------------------
health_problem_base <- process_glm_results(problem_models_base)
health_problem_control <- process_glm_results(problem_models_control)
health_protection_base <- process_glm_results(protection_models_base)
health_protection_control <- process_glm_results(protection_models_control)

# Changing to dataframe
health_problem_base <- as.data.frame(health_problem_base)
health_problem_control <- as.data.frame(health_problem_control)
health_protection_base <- as.data.frame(health_protection_base)
health_protection_control <- as.data.frame(health_protection_control)

# Assigning rownames
rownames(health_problem_base) <- health_problems
rownames(health_problem_control) <- health_problems
rownames(health_protection_base) <- health_protection
rownames(health_protection_control) <- health_protection

```

## Health Problems
### Base Model

```{r}
sjPlot::tab_model(
  problem_models_base[[1]]$model, problem_models_base[[2]]$model,
  problem_models_base[[3]]$model, problem_models_base[[4]]$model,
  problem_models_base[[5]]$model, problem_models_base[[6]]$model,
  problem_models_base[[7]]$model, problem_models_base[[8]]$model,
  problem_models_base[[9]]$model,
  pred.labels = c("(Intercept)", "Procrastination"),
  title = "Binary Logistic Regression Analysis Predicting Likelihood of Health Problems",
  dv.labels = health_problems, 
  collapse.ci = TRUE, p.style = "stars"
)
```

The results showed that procrastination was a significant predictor of all health problems except for having a heart condition, which had a non-significant positive association. The odds ratios ranged from 0.99 to 1.05, indicating small effects of procrastination on the odds of having a health problem.

```{r}
log_odds_plot(health_problem_base, title = "Risk of Experiencing a Health Problem")
```


### Control Model
```{r}
sjPlot::tab_model(
  problem_models_control[[1]]$model, problem_models_control[[2]]$model,
  problem_models_control[[3]]$model, problem_models_control[[4]]$model,
  problem_models_control[[5]]$model, problem_models_control[[6]]$model,
  problem_models_control[[7]]$model, problem_models_control[[8]]$model,
  problem_models_control[[9]]$model,
  pred.labels = c("Intercept", "Procrastination", "Depression", "Procrastination x Depression"),
  title = "Binary Logistic Regression Analysis Predicting Likelihood of Health Problems (with Control)",
  dv.labels = health_problems, 
  collapse.ci = TRUE, p.style = "stars")
```

After controlling for depression, the results showed that procrastination was a significant positive predictor of several health problems (back pain, fatigue, having high blood pressure, and diabetes)
```{r}
log_odds_plot(health_problem_control, title = "Risk of Experiencing a Health Problems (Controlling for Depression)")
```

## Health Protection
### Base Model
```{r}
sjPlot::tab_model(
  protection_models_base[[1]]$model, protection_models_base[[2]]$model,
  protection_models_base[[3]]$model, protection_models_base[[4]]$model,
  protection_models_base[[5]]$model, protection_models_base[[6]]$model,
  pred.labels = c("(Intercept)", "Procrastination"),
  dv.labels = health_protection, 
  title = "Binary Logistic Regression Analysis Predicting Likelihood of Health Protective Behaviours",
  collapse.ci = TRUE, p.style = "stars")
```

The results showed that procrastination was a significant negative predictor of several health protective behaviors (prostate exam, mammogram, cholesterol screening, and visiting the dentist). The odds ratios ranged from 0.98 to 0.99.

```{r}
log_odds_plot(health_protection_base, title = "Chance of Engaging in Health Protective Behaviours")
```

### Control Model
```{r}
sjPlot::tab_model(
  protection_models_control[[1]]$model, protection_models_control[[2]]$model,
  protection_models_control[[3]]$model, protection_models_control[[4]]$model,
  protection_models_control[[5]]$model, protection_models_control[[6]]$model,
  pred.labels = c("Intercept", "Procrastination", "Depression", "Procrastination x Depression"),
  title = "Binary Logistic Regression Analysis Predicting Likelihood of Health Protective Behaviours (with Control)",
  dv.labels = health_protection, 
  collapse.ci = TRUE, p.style = "stars")
```

After controlling for depression, the results showed that procrastination was a significant negative predictor of getting both a prostate and mammogram exam, and visiting the dentist. 

```{r}
log_odds_plot(health_protection_control, title = "Chance of Engaging in Health Protective Behaviours (Controlling for Depression)", size_font = 7)

```

