---
title: "Alcohol Analysis"
author: "Cormac Monaghan"
date: "`r format(Sys.time(), '%B %d, %Y')`"
toc: true
format:
  html:
    page-layout: full
    html-math-method: katex
    code-tools: true
    self-contained: true
    code-fold: true
    code-summary: "Show the code"
execute:
  warning: false
---
```{r setup, include = FALSE}
library(dplyr)
library(ggplot2)
library(visreg)
library(mgcv)

set.seed(1347) # Reproducibility

path_data <- "./01__Data/02__Processed/"
path_graphics <- "./02__Models/Results/Figures/03__GAM/"

health_data <- readxl::read_xlsx(here::here(path_data, "Health_HRS.xlsx"))

health_data <- health_data %>%
  rename(Days_drink = "Alcohol_week",
         Amount_drink = "Alcohol_amount") %>%
  mutate(Days_no_drink = 7 - Days_drink, .after = Days_drink)
```

# HRS (Alcohol Data)
The health and retirement study contains several questions related to alcohol and alcohol consumption.

- "Do you ever drink any alcoholic beverages such as beer, wine, or liquor."
- "In the last three months, on average, how many days per week have you had any alcohol to drink? - For example, beer, wine, or any drink containing liquor."
- "In the last three months, on the days you drink, about how many drinks do you have."

```{r alcohol-data}
health_data %>%
  select(ID, Alcohol, Days_drink, Days_no_drink, Amount_drink) %>%
  head(n = 5) %>%
  kableExtra::kbl(align = 'c') %>%
  kableExtra::kable_styling(
    full_width = FALSE,
    bootstrap_options = c("striped", "hover", "condensed", "responsive"))
  
```

# Exploratory Analysis
```{r alcohol-exploratory analysis, fig.show = 'hold', fig.ncol = 3}
# Do you drink alcohol
health_data %>%
  ggplot(aes(Alcohol)) +
  geom_bar(width = 0.5, fill = "skyblue", colour = "black") +
  stat_count(geom = "text", aes(label = ..count..), vjust = -0.5, size = 3) +
  scale_x_continuous(breaks = c(0, 1)) +
  ylim(0, 900) +
  labs(title = "Do you consume Alcohol") +
  theme_bw() +
  ggeasy::easy_center_title()

# How many days a week for you drink alcohol
health_data %>%
  ggplot(aes(Days_drink)) +
  geom_histogram(binwidth = 1, fill = "skyblue", colour = "black") +
  scale_x_continuous(breaks = seq(0, 7, by = 1)) +
  stat_count(geom = "text", aes(label = ..count..), vjust = -0.5, size = 3) +
  labs(title = "On average, how many days per week have you had any alcohol to drink?",
       x = "Days per week") +
  ylim(0, 900) +
  theme_bw() +
  ggeasy::easy_center_title()

# How much Alcohol per week do you drink
health_data %>%
  ggplot(aes(Amount_drink)) +
  geom_bar(width = 1, fill = "skyblue", colour = "black") +
  stat_count(geom = "text", aes(label = ..count..), vjust = -0.5, size = 3) +
  labs(title = " On the days you drink, about how many drinks do you have?",
       x = "# of drinks") +
  ylim(0, 900) +
  theme_bw() +
  ggeasy::easy_center_title()

# Procrastination
health_data %>%
  ggplot(aes(x = Days_drink / (Days_drink + Days_no_drink), y = Total_procrastination)) +
  geom_jitter(width = 0.05, alpha = 0.4) +
  labs(title = "Days Drank vs. Procrastination",
       x = "Proportion of Days Drank", y = "Total Procrastination") +
  theme_bw() +
  ggeasy::easy_center_title()

# Depression
health_data %>%
  ggplot(aes(x = Days_drink / (Days_drink + Days_no_drink), y = Total_depression)) +
  geom_jitter(width = 0.05, alpha = 0.4) +
  labs(title = "Days Drank vs. Depression",
       x = "Proportion of Days Drank", y = "Total Depression") +
  theme_bw() +
  ggeasy::easy_center_title()

# Age
health_data %>%
  ggplot(aes(x = Days_drink / (Days_drink + Days_no_drink), y = Age)) +
  geom_jitter(width = 0.05, alpha = 0.4) +
  labs(title = "Days Drank vs. Age",
       x = "Proportion of Days Drank", y = "Age") +
  theme_bw() +
  ggeasy::easy_center_title()
```

# Modelling
## Binomial GLM (one predictor)

**Overview**

$$
log \; \frac{p_i}{1-p_i} = \beta_0 + \beta_1x_i
$$
where $p_i$ is the probability of drinking on a given day for the $i$-th person, $x_i$ is the total procrastination score for the $i$-th person, and $\beta_0$ and $\beta_1$ are the coefficients to be estimated.

**Model Fitting**
```{r glm_1_fit}
fit_1 <- glm(cbind(Days_drink, Days_no_drink) ~ Total_procrastination,
           data = health_data, family = binomial(link = "logit"))

summary(fit_1)
```

**Diagnostics**
```{r glm_1_diagnostics}
# Using a half normal plot with simulated envelopes
hnp::hnp(fit_1, main = "HNP for binomial GLM")
```

**Visualizing**
```{r glm_1_visualisation}
visreg(fit = fit_1, 
       xvar = "Total_procrastination", 
       gg = TRUE, 
       scale = "response", 
       rug = FALSE) +
  geom_jitter(data = health_data, aes(x = Total_procrastination, y = Days_drink / 7), 
              alpha = 0.5, size = 0.9) +
  scale_x_continuous(breaks = seq(0, 60, by = 10)) +
  labs(title = "Relationship between days drinking and procrastination (GLM)",
       x = "Total Procrasination", 
       y = "prob(Days Drinking)") +
  theme_bw() +
  ggeasy::easy_center_title()
```

## Binomial GLM (multiple predictors)

**Overview**

$$
log \; \frac{p_i}{1-p_i} = \beta_0 + \beta_1x_{1i} + \beta_2x_{2i} + \dots + \beta_px_{qi}
$$

where $p_i$ is the probability of drinking on a given day for the $i$-th person, $\beta_0, \beta_1, \dots, \beta_p$ are the coefficients to be estimated for each independent variable, and $x_{1i}, x_{2i}, \dots, x_{qi}$ are the independent variables themselves for the $i$-th person.

**Model Fitting**
```{r glm_2_fit}
fit_2 <- glm(cbind(Days_drink, Days_no_drink) ~ Total_procrastination + Total_depression + Education + Age, 
             data = health_data, family = binomial(link = "logit"))

summary(fit_2)
```

**Diagnostics**
```{r glm_2_diagnostics}
# Using a half normal plot with simulated envelopes
hnp::hnp(fit_2, main = "HNP for binomial GLM")
```

**Visualizing**
```{r glm_2_visualisation}
visreg(fit = fit_2, 
       xvar = "Total_procrastination", 
       gg = TRUE, 
       scale = "response", 
       rug = FALSE) +
  geom_jitter(data = health_data, aes(x = Total_procrastination, y = Days_drink / 7), 
              alpha = 0.5, size = 0.9) +
  scale_x_continuous(breaks = seq(0, 60, by = 10)) +
  labs(title = "Relationship between days drinking and procrastination (GLM)",
       subtitle = "Controlling for depression, age, and education",
       x = "Total Procrasination", y = "prob(Days Drinking)") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

## Binomial GAM

**Overview**
$$
log \; \frac{p_i}{1 - p_i} = \alpha + f_1(x_{1i}) + f_2(x_{2i}) + \dots + f_q(x_{qi})
$$
where $p_i$ is the probability of drinking on a given day, $\alpha$ is the intercept term, $f_1, \dots, f_q$ are smooth function of the predictor variables $x_{1i}, \dots, x_{qi}$. The smooth functions $f_q$ can be estimated by various methods, such as splines, kernels, or local regression.

**Model Fitting**
```{r gam_fit}
fit_3 <- gam(cbind(Days_drink, Days_no_drink) ~ s(Total_procrastination) + s(Total_depression, k = 9) + s(Age) + Education, 
             data = health_data, family = "binomial", method = "REML")

summary(fit_3)
```

**Diagnostics**
```{r gam_residuals}
fit_3_residuals <- DHARMa::simulateResiduals(fit_3)
plot(fit_3_residuals) 
```

**Visualization**
```{r, fig.show = 'hold'}
plot(fit_3, select = 1, trans = plogis, shift = fit_3$coefficients[1], seWithMean = TRUE, 
     rug = FALSE, shade = TRUE, shade.col = "skyblue",
     ylab = "Probability", xlab = "Total Procrastination",
     main = "Probability of drinking on a given day by procrastination")

plot(fit_3, select = 2, trans = plogis, shift = fit_3$coefficients[1], seWithMean = TRUE, 
     rug = FALSE, shade = TRUE, shade.col = "skyblue",
     ylab = "Probability", xlab = "Total Depression",
     main = "Probability of drinking on a given day by depression")

plot(fit_3, select = 3, trans = plogis, shift = fit_3$coefficients[1], seWithMean = TRUE, 
     rug = FALSE, shade = TRUE, shade.col = "skyblue",
     ylab = "Probability", xlab = "Age",
     main = "Probability of drinking on a given day by age")

# Plotting using visreg
visreg(fit = fit_3, xvar = "Total_procrastination", gg = TRUE, scale = "response", rug = FALSE) + 
  geom_jitter(data = health_data, aes(x = Total_procrastination, y = Days_drink / 7), 
              alpha = 0.5) +
  scale_x_continuous(breaks = seq(0, 60, by = 10)) +
  labs(title = "Relationship between days drinking and procrastination (GAM)",
       subtitle = "Controlling for depression, age, and education",
       x = "Total Procrasination", y = "prob(Days Drinking)") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))
```

